---
title: "Assignment 5: Data Visualization"
author: "Jackie Van Der Hout"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(cowplot)
NTL_LTER <- read.csv("../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
NEON_LITTER <- read.csv("../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2
class(NTL_LTER$sampledate)
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%Y-%m-%d")
class(NEON_LITTER$collectDate)
NEON_LITTER$collectDate <- as.Date(NEON_LITTER$collectDate, format = "%Y-%m-%d")
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_light(base_size = 13) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
tp_ugXpo4 <- ggplot(NTL_LTER, aes(x = po4, y = tp_ug, color = lakename))+
  geom_point(alpha = 0.5)+
  labs(x = "Total Phosphate", y = "Total Phospohorous", title = "Phosophorous and Phosphate in Peter and Paul Lakes", color = "Lake")+
  xlim(0,50)+
  ylim(0, 150)+
  geom_smooth(method = lm, color = "black", se = FALSE)+
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), legend.title = element_text(size = 12))
tp_ugXpo4
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r, fig.height = 7, fig.width = 6}
#5
library(gridGraphics)

NTL_LTER$month <- as.factor(NTL_LTER$month)
class(NTL_LTER$month)

tempbox <-ggplot(NTL_LTER, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot()+
  labs(x = "Month", y = "Temperature", color = "Lake")+
  theme(legend.position = "none")
tempbox   

TPbox <- ggplot(NTL_LTER, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot()+
  labs(x = "Month", y = "Total Phosphorous", color = "Lake")+
  theme(legend.position = "none")
TPbox

TNbox <- ggplot(NTL_LTER, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot()+
  labs(x = "Month", y = "Total Nitrogen", color = "Lake")+
  theme(legend.position = "none")
TNbox   

boxes <- plot_grid(
  tempbox + theme(legend.position = "none"),
  TPbox + theme(legend.position = "none"),
  TNbox + theme(legend.position = "none"),
  align = 'vh',
  hjust = -1,
  nrow = 1
)

fakelegend <- get_legend(tempbox +
  theme(legend.position = "right", legend.box.margin = margin(0, 0, 0, 1))
  )

plot_grid(boxes, fakelegend, ncol = 2, rel_widths = c(3, .4))
#this seems like a clunky way to do it but it more or less works
#however the output plot only looks good when expanded to full size 
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature has a wider range of variability than Phosphorous and Nitrogen, and is also the most sensitive to seasonal changes in both lakes. Both Peter and Paul lakes appear to display similar seasonal changes in temperature, with similar standard error ranges. However, Paul Lake seems to display a slightly higher mean temperature than Peter Lake until October, when Peter lake cools down more slowly. Inversely, Peter Lake appears to have a higher concentration of both Phosphorous and Nitrogen on average, as well as higher outlier peak values in nutrient concentration than Paul Lake. Nutrient concentrations appear to be seasonally highest in both lakes and in both nutrient categories during the late summer / early fall. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
needleplot <- ggplot(subset(NEON_LITTER, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass, color = nlcdClass))+
  geom_point()+
  labs(x = "Date", y = "Litter Dry Mass", color = "Land Cover Type")+
  theme(legend.position = "right")
needleplot

#7
NEON_LITTER$nlcdClass <- factor(NEON_LITTER$nlcdClass, levels = c("evergreenForest", "grasslandHerbaceous", "shrubScrub"), 
                  labels = c("Evergreen Forest", "Herbacious Grassland", "Shrublands"))

needleplot2 <- ggplot(subset(NEON_LITTER, functionalGroup == "Needles"), aes(x = collectDate, y = dryMass))+
  geom_point()+
  labs(x = "Date", y = "Litter Dry Mass")+
  theme(legend.position = "right")+
  facet_wrap(vars(nlcdClass), nrow = 1)
needleplot2
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: The facet wrapped plot in #7 provides an easier comparison of litter dry mass over both land cover type and time because of the side-by-side comparison with the same axes, instead of having all of the datapoints within one plot. 
