---
title: "Assignment 09: Data Scraping"
author: "Jackie Van Der Hout"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(rvest)
library(lubridate)
theme_set(theme_light())
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Max Daily Use (MGD) - for each month 


In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- webpage %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text
water.system.name

pswid <- webpage %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text
pswid

ownership <- webpage %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text
ownership

max.withdrawals.mgd <- webpage %>% html_nodes("th~ td+ td") %>% html_text
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4
df_withdrawals <- data.frame("Month" = rep(1:12),
                             "Year" = (2020),
                             "WaterSystem" = water.system.name,
                             "pswid" = pswid,
                             "Ownership" = ownership,
                             "MaxWithdrawlMGD" = as.numeric(max.withdrawals.mgd)) %>% 
  mutate(Date = my(paste(Month,"-",Year)))


#5
ggplot(df_withdrawals, aes(x=Date, y= MaxWithdrawlMGD)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2020 Water usage data for", water.system.name),
       subtitle = ownership,
       y="Withdrawal (mgd)",
       x="Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6.
the_year <- 2020 

scrape.it <- function(pswid, the_year){
  
  #Retrieve the website contents 
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?' , 
                                  'pwsid=', pswid, '&year=', the_year))
  
  #Set the element address variables (determined in the previous step)
  water.system.name_tag <- '.table tr:nth-child(1) td:nth-child(2)'
  pswid_tag <- 'tr:nth-child(2) th+ .left:nth-child(2)'
  ownership_tag <- 'tr:nth-child(2) .left~ .left+ td.left'
  max.withdrawals.mgd_tag <- '.table:nth-child(7) td:nth-child(7) , .table:nth-child(7) td:nth-child(3)'
  
  #Scrape the data items
  water.system.name.scrape <- the_website %>% html_nodes(water.system.name_tag) %>% html_text()
  pswid.scrape <- the_website %>%   html_nodes(pswid_tag) %>%  html_text()
  ownership.scrape <- the_website %>% html_nodes(ownership_tag) %>% html_text()
  withrdrawls.scrape <- the_website %>% html_nodes(max.withdrawals.mgd_tag) %>% html_text()
  
  #Convert to a dataframe
  df_withdrawals.scraped <- data.frame("Month" = rep(1:12),
                             "Year" = rep(the_year, 12),
                             "WaterSystem" = as.character(water.system.name.scrape),
                             "pswid" = pswid.scrape,
                             "Ownership" = ownership.scrape,
                             "MaxWithdrawlMGD" = as.numeric(withrdrawls.scrape)) %>% 
  mutate(Date = my(paste(Month,"-",Year)))
  
  return(df_withdrawals.scraped)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
Durham_2015 <- scrape.it('03-32-010',2015)
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8a
Asheville_2015 <- scrape.it('01-11-010',2015)
view(Asheville_2015)

#8b
Withdrawls2015 <- bind_rows(Durham_2015, Asheville_2015)

#8c
ggplot(Withdrawls2015,aes(x=Date, y= MaxWithdrawlMGD, color = WaterSystem)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Municipal Water Withdrawls in 2015", subtitle = "Durham and Asheville", y= "Withdrawal (mgd)", x="Date", color = "City")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9a
the_years = rep(2010:2019)

Asheville10_19 <- the_years %>% 
  map(scrape.it, pswid = '01-11-010') %>% 
  bind_rows()
view(Asheville10_19)

#9b
ggplot(Asheville10_19,aes(x=Date, y= MaxWithdrawlMGD)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Asheville Municipal Water Withdrawls", subtitle = "2010 - 2019", y= "Withdrawal (mgd)", x="Date", color = "City")
```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
Yes, there seems to be a sharp increase in water usage in Asheville over time. 